---
title: "The mixing-period tag treatment in the 2026 WCPO bigeye assessment"
subtitle: "What `tag_flags(i,2)` can and cannot do, read against `bet.tag` and the fitted reporting rates"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    embed-resources: true
    code-fold: true
    code-tools: true
    df-print: kable
params:
  source_mode: "remote_cache"
  tag_path: null
  par_path: null
  cache_dir: "inputs/cache"
  refresh_inputs: false
  source_ref: "8c2a586a8c5abec1a76e305f28f8f0b48fe2fe27"
---

```{r setup}
#| include: false
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      collapse = TRUE, comment = "#>")

if (!exists("params")) {
  params <- list(source_mode = "remote_cache", tag_path = NULL, par_path = NULL,
                 cache_dir = "inputs/cache", refresh_inputs = FALSE,
                 source_ref = "8c2a586a8c5abec1a76e305f28f8f0b48fe2fe27")
}

repo_raw_url <- function(ref, path) {
  paste0("https://raw.githubusercontent.com/PacificCommunity/",
         "ofp-sam-bet-2026-diagnostic/", ref, "/", path)
}

acquire_input <- function(name, remote_path, local_path = NULL) {
  if (!is.null(local_path) && nzchar(local_path)) {
    if (!file.exists(local_path)) stop("Local override does not exist: ", local_path)
    return(list(path = normalizePath(local_path), source = "local override"))
  }
  if (!identical(params$source_mode, "remote_cache")) {
    stop("source_mode must be 'remote_cache' unless an explicit local path is supplied")
  }

  cache_path <- file.path(params$cache_dir, params$source_ref, name)
  if (isTRUE(params$refresh_inputs) || !file.exists(cache_path)) {
    dir.create(dirname(cache_path), recursive = TRUE, showWarnings = FALSE)
    temporary_path <- paste0(cache_path, ".download")
    unlink(temporary_path)
    status <- tryCatch(
      download.file(repo_raw_url(params$source_ref, remote_path), temporary_path,
                    mode = "wb", quiet = TRUE),
      error = function(error) error
    )
    if (inherits(status, "error") || !identical(status, 0L) ||
        !file.exists(temporary_path) || file.info(temporary_path)$size == 0) {
      unlink(temporary_path)
      stop("Could not download ", remote_path, " at ref ", params$source_ref,
           ". Re-render after restoring network access or provide an explicit local path.")
    }
    backup_path <- paste0(cache_path, ".previous")
    unlink(backup_path)
    if (file.exists(cache_path) && !file.rename(cache_path, backup_path)) {
      unlink(temporary_path)
      stop("Could not preserve existing cached input: ", cache_path)
    }
    if (!file.rename(temporary_path, cache_path)) {
      if (file.exists(backup_path)) file.rename(backup_path, cache_path)
      unlink(temporary_path)
      stop("Could not move downloaded input into cache: ", cache_path)
    }
    unlink(backup_path)
    return(list(path = normalizePath(cache_path), source = "downloaded"))
  }
  list(path = normalizePath(cache_path), source = "cache")
}

input_files <- list(
  tag = acquire_input("bet.tag", "model/bet.tag", params$tag_path),
  par = acquire_input("final.par", "results/reference/final.par", params$par_path)
)
stopifnot(file.exists(input_files$tag$path), file.exists(input_files$par$path))

# Formatting helpers, matching the conventions of the companion simulation study.
fpc  <- function(x, d = 1) sprintf(paste0("%.", d, "f%%"), 100 * x)
fnum <- function(x) formatC(x, big.mark = ",", format = "d")
```

::: {.callout-note}
## Scope

This note is an executive summary of a methodological problem in the 2026 bigeye
diagnostic model: the treatment of tag recaptures during the mixing period, controlled
by `tag_flags(i,2)`. It computes, directly from `bet.tag` and the model's flag settings,
the two quantities that govern the problem — the share of recapture data excluded from
the likelihood, and the mixing-period recapture fraction $\hat\psi$ raised by the fitted
fishery-specific reporting rates — and reads them against a companion simulation study
(*Mixing-period tag recaptures in MULTIFAN-CL*, v11) whose results are quoted with
provenance, not re-derived.

Everything computed here is parsed inline from the `.tag` and fitted `.par` inputs;
everything quoted from the simulation study or reported from the assessment's
uncertainty grid is labelled as such. The recommendation at the end — a mixing-period
likelihood as the resolution, and a tag-free fleets-as-areas configuration as the
practical bound — follows from the demonstrated deficiencies of both flag settings, not
from any simulation of a spatial model, which no document in this series performs.
:::

::: {.callout-note}
## Input provenance

Model inputs are pinned to
[`ofp-sam-bet-2026-diagnostic`](https://github.com/PacificCommunity/ofp-sam-bet-2026-diagnostic)
commit `r params$source_ref`. The tag file was resolved from
`r input_files$tag$source` at `r input_files$tag$path`; the fitted parameter file was
resolved from `r input_files$par$source` at `r input_files$par$path`.

When no local overrides are supplied, the document reads
`model/bet.tag` and `results/reference/final.par` from the pinned GitHub revision and
caches them under `r params$cache_dir`.
:::

::: {.callout-warning}
## Disclaimer: use of AI assistance

This note was prepared with the assistance of a large language model (Anthropic Claude).
The model parsed `bet.tag`, transcribed the flag settings and the `tag fish rep`
reporting-rate matrix from the supplied diagnostic `.par`, computed the diagnostics, and
drafted the text.

Output produced this way requires independent verification, and the document is
structured so that verification is possible rather than requiring trust:

- all computation is shown inline and runs in base R with no package dependencies;
- structural tripwires (`stopifnot`) check the parse against the file's own recovery
  header before any result is used, so a parsing failure halts the render rather than
  producing wrong numbers;
- the R implementation mirrors an independently written and executed implementation;
  any divergence between the two will surface as a tripwire failure or as numbers
  differing from the companion diagnostic note (v2), which reports the same quantities;
- the per-group appendix permits any single row to be checked by hand against the
  `.tag` file in a few minutes;
- every value quoted from the simulation study is attributed to it, and every value
  reported from the assessment's uncertainty grid is labelled as reported, not verified.

The inputs are read from a pinned upstream model revision or explicit local overrides.
The parser's section and dimension tripwires should still be checked before relying on
the document. Conclusions and recommendations are the responsibility of the author,
not of the tool.
:::

```{r constants}
#| code-summary: "Parse fitted tag settings and define quoted comparison inputs"

read_numeric_section <- function(path, marker, n_columns) {
  lines <- gsub("\r$", "", readLines(path, warn = FALSE))
  marker_at <- grep(paste0("^#\\s*", marker, "\\s*$"), lines,
                    ignore.case = TRUE)
  if (length(marker_at) != 1) stop("Expected one '# ", marker, "' section in ", path)
  next_marker <- grep("^#", lines[(marker_at + 1):length(lines)])[1]
  end_at <- if (is.na(next_marker)) length(lines) else marker_at + next_marker - 1
  rows <- trimws(lines[(marker_at + 1):end_at])
  rows <- rows[nzchar(rows)]
  fields <- strsplit(rows, "\\s+")
  if (!length(fields) || any(lengths(fields) != n_columns)) {
    stop("Malformed '# ", marker, "' section in ", path)
  }
  values <- suppressWarnings(as.numeric(unlist(fields, use.names = FALSE)))
  if (anyNA(values)) stop("Non-numeric value in '# ", marker, "' section in ", path)
  matrix(values, ncol = n_columns, byrow = TRUE)
}

tag_flags <- read_numeric_section(input_files$par$path, "tag flags", 10)
RR <- read_numeric_section(input_files$par$path, "tag fish rep", 33)
nmix <- tag_flags[, 1]

stopifnot(
  nrow(tag_flags) == 98,
  nrow(RR) >= 98,
  length(nmix) == 98,
  all(tag_flags[, 2] == 1),
  ncol(RR) == 33
)

RR <- RR[seq_len(98), , drop = FALSE]

# Regional seeding-prior means (SC22-SA-IP-05, purse seine), used only for comparison.
X_reg <- c(`1` = 0.5, `2` = 0.496, `3` = 0.512, `4` = 0.528, `5` = 0.5)

# --- Quoted from the companion simulation study (v11 fitted objects), not re-derived. ---
sim <- list(
  expD = data.frame(psi  = c(0.05, 0.30, 0.50),                    # Experiment D:
                    best = c(-0.039, -0.277, -0.509),              # bias in mixing-period F
                    worst= c(-0.048, -0.334, -0.599)),             # worst end = longest mixing
  floor_tau   = 0.22,     # survival floor; engages when removal fraction > 1 - tau = 0.78
  expE_N      = 125,      # Experiment E small-group design point
  expE_tail   = 0.060,    # share of groups with F-hat > 5x truth at that size
  expE_tail_big = 0.000,  # the same at 10,000 tags per group
  flag_gap_max  = 0.63)   # largest flag0-vs-truth gap in design (phi = 2, nmix = 4)
```

```{r parse}
#| code-summary: "Parse bet.tag (base R) with structural tripwires"

lines <- gsub("\r$", "", readLines(input_files$tag$path))

hdr <- as.integer(strsplit(trimws(lines[2]), "\\s+")[[1]])
n_groups <- hdr[1]; n_bins <- hdr[3]
stopifnot(n_groups == 98)

rec_hdr_at <- grep("^# TAG RECOVERIES", lines)[1]
totals <- as.integer(strsplit(trimws(lines[rec_hdr_at + 2]), "\\s+")[[1]])
stopifnot(length(totals) == n_groups)

blk <- grep("^#\\s+[0-9]+\\s+-\\s+RELEASE REGION", lines)
stopifnot(length(blk) == n_groups)
ends <- c(blk[-1] - 1, length(lines))

groups <- vector("list", n_groups)
recs   <- vector("list", n_groups)
for (g in seq_len(n_groups)) {
  seg  <- lines[blk[g]:ends[g]]
  prog <- tail(strsplit(trimws(seg[1]), "\\s+")[[1]], 1)
  body <- seg[-1]
  keep <- !grepl("^\\s*#", body) & nzchar(trimws(body))
  body <- trimws(body[keep])
  relh <- as.integer(strsplit(body[1], "\\s+")[[1]])          # region, year, month
  rel  <- as.numeric(strsplit(body[2], "\\s+")[[1]])
  stopifnot(length(rel) == n_bins)
  rmat <- if (length(body) > 2) {
    do.call(rbind, lapply(strsplit(body[-(1:2)], "\\s+"), as.integer))
  } else matrix(integer(0), 0, 5)
  stopifnot(nrow(rmat) == totals[g])   # header row counts recovery records per group
  q <- if (nrow(rmat)) ((rmat[, 3] - relh[2]) * 12 + (rmat[, 4] - relh[3])) %/% 3L
       else integer(0)
  stopifnot(all(q >= 0), all(((rmat[, 3] - relh[2]) * 12 + (rmat[, 4] - relh[3])) %% 3 == 0))
  groups[[g]] <- data.frame(gid = g, prog = prog, region = relh[1], year = relh[2],
                            N = sum(rel))
  recs[[g]]   <- if (nrow(rmat)) data.frame(gid = g, q = q, fishery = rmat[, 2],
                                            n = rmat[, 5]) else NULL
}
grp <- do.call(rbind, groups)
rec <- do.call(rbind, recs)
stopifnot(identical(grp$prog, c(rep("RTTP", 15), rep("PTTP", 46), rep("JPTP", 37))))
stopifnot(n_groups == nrow(tag_flags), all(rec$fishery >= 1),
          all(rec$fishery <= ncol(RR)))

grp$nmix <- nmix
rec$rate <- RR[cbind(rec$gid, rec$fishery)]
rec$mix  <- rec$q < grp$nmix[rec$gid]

N_tot <- sum(grp$N); R_tot <- sum(rec$n)
# Note: the recovery header row counts records, not tags; it is checked per group above.
```

```{r compute}
#| code-summary: "The two governing quantities"

# 1. Share of recapture data excluded from the likelihood by the mixing windows.
R_mix   <- sum(rec$n[rec$mix])
mix_share <- R_mix / R_tot

# 2. psi-hat per group: mixing returns raised by the fitted rate of the reporting
#    fishery (zero-rate returns flagged and excluded), divided by adjusted releases.
zero_mix <- sum(rec$n[rec$mix & rec$rate == 0])
raised <- with(rec[rec$mix & rec$rate > 0, ], tapply(n / rate, gid, sum))
grp$Rmix <- as.integer(with(rec[rec$mix, , drop = FALSE],
                            tapply(n, factor(gid, levels = 1:98), sum)))
grp$Rmix[is.na(grp$Rmix)] <- 0L
grp$psi <- 0; grp$psi[as.integer(names(raised))] <- raised
grp$psi <- grp$psi / grp$N
grp$psi_reg <- grp$Rmix / (X_reg[as.character(grp$region)] * grp$N)
grp$R <- with(rec, tapply(n, factor(gid, levels = 1:98), sum)); grp$R[is.na(grp$R)] <- 0

w_psi     <- sum(pmin(grp$psi, 1)     * grp$N) / N_tot
w_psi_reg <- sum(pmin(grp$psi_reg, 1) * grp$N) / N_tot
w_by_prog <- sapply(split(grp, grp$prog), function(d)
  sum(pmin(d$psi, 1) * d$N) / sum(d$N))

hi <- grp$psi >= 0.30
rec_hi <- sum(grp$R[hi]) / R_tot

# Switching flag=1 -> flag=0: extra raised removals by rate band.
band <- cut(rec$rate, c(-1, 0, 0.4, 0.95, 1.01),
            labels = c("zero-rate", "longline 0.16-0.26", "purse seine 0.46-0.84",
                       "JP coastal 0.99"))
extra <- with(rec[rec$mix & rec$rate > 0, ],
              tapply(n / rate - n, band[rec$mix & rec$rate > 0], sum))
extra[is.na(extra)] <- 0

# Counterfactual flag=0 floor margins under fitted rates and configured windows.
floor_util <- sapply(seq_len(98), function(g) {
  d <- rec[rec$gid == g & rec$mix & rec$rate > 0, , drop = FALSE]
  N <- grp$N[g]; u <- 0
  for (t in 0:(max(grp$nmix[g], 1) - 1)) {
    rt <- sum(d$n[d$q == t] / d$rate[d$q == t])
    if (N <= 0) return(Inf)
    u <- max(u, (rt / N) / (1 - sim$floor_tau)); N <- N - rt
  }
  u
})
n_floor <- sum(floor_util >= 1)
```

# Summary {#sec-summary}

**Two structural facts frame everything.** First, under the configured mixing windows,
`r fnum(R_mix)` of the `r fnum(R_tot)` reported recaptures in `bet.tag` —
**`r fpc(mix_share)` of the tagging data** — never enter the likelihood. Under either
setting of `tag_flags(i,2)`, those returns fix the tagged population state
deterministically, with full authority and no uncertainty attached, and contribute
nothing to parameter estimation. Second, the removal that fixes that state depends
entirely on what is assumed about reporting rates during mixing, and the model's own
fitted `tag fish rep` matrix says those rates are far from uniform: near-complete
(0.99) for the Japanese coastal fisheries that dominate JPTP returns, roughly one half
(0.46--0.50) for the main purse-seine fisheries that dominate PTTP and RTTP returns,
and 0.16--0.26 for longline.

**The configured setting, `tag_flags(i,2) = 1`, removes reported mixing-period returns
at face value — equivalent to assuming complete reporting — and its cost falls almost
entirely on the purse-seine programmes.** By the model's own fitted rates the
assumption is nearly true for JPTP (weighted $\hat\psi$ = `r round(w_by_prog["JPTP"], 3)`;
`r fpc(sum(rec$n[rec$mix & rec$fishery %in% c(12,13) & grp$prog[rec$gid] == "JPTP"]) / sum(rec$n[rec$mix & grp$prog[rec$gid] == "JPTP"]), 0)`
of JPTP mixing returns come from the 0.99-rate fisheries) and substantially false for
RTTP (`r round(w_by_prog["RTTP"], 3)`) and PTTP (`r round(w_by_prog["PTTP"], 3)`), where
roughly a third to a half of true mixing recaptures are never removed. The companion
study's Experiment D maps $\hat\psi$ to the downward bias in mixing-period $F$:
`r fpc(abs(sim$expD$best[1]), 1)`--`r fpc(abs(sim$expD$worst[1]), 1)` at $\psi$ = 0.05,
`r fpc(abs(sim$expD$best[2]), 1)`--`r fpc(abs(sim$expD$worst[2]), 1)` at 0.3, and
`r fpc(abs(sim$expD$best[3]), 1)`--`r fpc(abs(sim$expD$worst[3]), 1)` at 0.5, the worse
end at longer mixing. `r sum(hi)` release groups sit at $\hat\psi \ge 0.3$, carrying
`r fpc(rec_hi, 0)` of the reported recaptures. The bias is systematic — $F$ low, scale
high, every group the same direction — and its one merit is that it is predictable in
advance from exactly this calculation.

**The alternative, `tag_flags(i,2) = 0`, raises returns by the assumed rate before
removal, and is unbiased only under conditions this file does not fully meet.** Its
accuracy moves one-for-one with the reporting rates, which are the quantity under
dispute, weakly informed for longline, and being revised downward. Its inversion is
convex in a noisy observation, so small release groups put weight in the upper tail of
$F$: at the study's `r sim$expE_N`-tag design point, `r fpc(sim$expE_tail, 1)` of
simulated groups returned estimates more than five times the truth with everything else
correct, invisible to median-based diagnostics — and `r sum(grp$N <= sim$expE_N)` of the
98 groups here are that small or smaller. The oft-cited floor pathology, by contrast,
largely dissolves under the fitted rates: re-evaluated with fishery-specific raising and
the configured windows, `r n_floor` group(s) breach the survival-floor condition, not
the six that regional-mean raising suggests. And the switch is surgically targeted:
moving to raised removals adds `r fnum(round(sum(extra[c("purse seine 0.46-0.84","longline 0.16-0.26")])))`
tags to the purse-seine and longline removals — where the face-value bias lives — while
changing JPTP by `r fnum(round(extra["JP coastal 0.99"]))` tags in total.

**Running both settings as levels of the uncertainty grid does not bound the problem.**
The pair is structurally one-sided: because the assumed rate is below one wherever it
matters, the raised treatment always removes more tags and always returns the larger
$F$. Two grid levels therefore span an assumption without bracketing the truth — they
cannot contain it at all when the assumed rate is too high — and, because one level
carries a documented directional bias, averaging them functions as a fifty per cent
weighting on that bias, not as an uncertainty representation.

**This is not academic: the choice reportedly flips the status determination.** With
both settings in the grid, the ensemble indicates the stock is not overfished and not
undergoing overfishing; restricted to `tag_flags(i,2) = 0`, it indicates overfished and
about as likely as not undergoing overfishing. Those ensemble results are reported here,
not verified, but their direction is exactly what the $\hat\psi$ analysis predicts, and
the guidance against the grid axis was fixed before the ensemble outcome was known.
The honest statement of status is therefore not either determination — it is that the
determination currently turns on a model component that is biased under both of its
available settings.

**The resolution and the way forward are different things.** The resolution is a
mixing-period likelihood: predict mixing-period returns, score them against
observations, and estimate the reporting rate against its tag-seeding prior — putting
`r fpc(mix_share, 0)` of the tag data back into estimation and inverting nothing. The
companion study implements this (`forward_X`) and it recovers $F$ under the
misspecification that biases both flag settings; MULTIFAN-CL cannot currently do it, so
it is a redevelopment request, not a configuration, and the study also shows it is not
free with many small release groups. Until it exists, the most defensible way forward is
to bound the component rather than tune it: include a tag-free configuration, with
spatial structure carried by a fleets-as-areas approach, in the structural uncertainty
grid. Given that the status determination demonstrably moves with the tag treatment, the
question that configuration answers — how much of the assessment result depends on the
tagging data at all — is now the question managers most need answered.

# The $\hat\psi$ table {#sec-psi}

$\hat\psi_i$ raises each mixing-window return by the fitted rate of the fishery that
reported it, then divides by the correction-adjusted release count. Omitting the raising
returns $X\psi$ rather than $\psi$ — roughly half the diagnosed fraction at purse-seine
rates, and because the bias steepens in $\psi$, less than half the diagnosed
consequence. `r fnum(zero_mix)` mixing returns come from a fishery whose fitted rate is
zero and cannot be raised at all; they are excluded, which understates $\hat\psi$ for
the affected groups, and the inconsistency should be resolved regardless of the flag
choice.

```{r psi-table}
#| echo: false
cls <- cut(grp$psi, c(-1, 0.05, 0.30, 0.50, Inf),
           labels = c("< 0.05", "0.05\u20130.30", "0.30\u20130.50", "\u2265 0.50"))
tab <- data.frame(
  `psi class` = levels(cls),
  Groups = as.integer(table(cls)),
  `Share of recaptures` = sapply(levels(cls), function(k)
    fpc(sum(grp$R[cls == k]) / R_tot, 1)),
  `Experiment D bias class (mixing-period F)` =
    c("a few per cent", "up to roughly one third",
      "roughly one third to one half", "one half or more"),
  check.names = FALSE)
knitr::kable(tab, row.names = FALSE)

byprog <- do.call(rbind, lapply(split(grp, grp$prog), function(d) data.frame(
  Programme = d$prog[1], Groups = nrow(d), Releases = fnum(round(sum(d$N))),
  `Mixing returns` = fnum(sum(d$Rmix)),
  `Weighted psi (fishery rates)` = round(sum(pmin(d$psi, 1) * d$N) / sum(d$N), 3),
  `(regional means)` = round(sum(pmin(d$psi_reg, 1) * d$N) / sum(d$N), 3),
  check.names = FALSE)))
knitr::kable(byprog[order(match(byprog$Programme, c("RTTP", "PTTP", "JPTP"))), ],
             row.names = FALSE)
```

The programme contrast carries the story: the face-value assumption is nearly correct
where the fitted rates are near one, and costs a third to a half of the true mixing
removals where they are near one half — which is where
`r fpc(sum(grp$N[grp$prog != "JPTP"]) / N_tot, 0)` of the releases sit.

# Limitations

1. **The transcribed settings are the load-bearing inputs.** `nmix` and the
   reporting-rate patterns must be verified against the final `.par` before this note
   is relied on; a render against different settings will silently produce different —
   correct — numbers.
2. **The fitted rates were estimated under `tag_flags(i,2) = 1`.** Using them to
   characterise the raised counterfactual is indicative; under the default the rates
   would re-equilibrate, and the feedback runs toward the floor. The one-group floor
   finding should be confirmed from an actual `tag_flags(i,2) = 0` fit before it is
   quoted without qualification.
3. **The Experiment D mapping is from a deliberately stripped-down operating model.**
   It gives the class of the bias, not its value in a full assessment; the companion
   study's limitations section applies in full.
4. **The ensemble status results are reported, not reproduced.** This note takes the
   direction and existence of the flip as given by the assessment team; nothing here
   recomputes grid outputs.
5. **$\hat\psi$ inherits the rates it is raised by**, in the direction that understates
   the problem if reporting estimates fall further, and the zero-rate fishery-27
   returns are excluded from raising.

# Appendix: per-group $\hat\psi$ {#sec-appendix}

```{r appendix}
#| echo: false
app <- data.frame(Group = grp$gid, Prog. = grp$prog, Region = grp$region,
                  Year = grp$year, `tag_flags(i,1)` = grp$nmix,
                  N = round(grp$N), R = grp$R, `R mix` = grp$Rmix,
                  `psi (fishery rates)` = round(grp$psi, 3),
                  `psi (regional means)` = round(grp$psi_reg, 3),
                  check.names = FALSE)
knitr::kable(app, row.names = FALSE)
```
