---
title: "The mixing-period tag treatment in the 2026 WCPO bigeye assessment"
subtitle: "What `tag_flags(i,2)` can and cannot do, read against `bet.tag` and the fitted reporting rates"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    embed-resources: true
    code-fold: true
    code-tools: true
    df-print: kable
params:
  source_mode: "remote_cache"
  tag_path: null
  par_path: null
  cache_dir: "inputs/cache"
  refresh_inputs: false
  source_ref: "8c2a586a8c5abec1a76e305f28f8f0b48fe2fe27"
---

```{r setup}
#| include: false
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      collapse = TRUE, comment = "#>")

if (!exists("params")) {
  params <- list(source_mode = "remote_cache", tag_path = NULL, par_path = NULL,
                 cache_dir = "inputs/cache", refresh_inputs = FALSE,
                 source_ref = "8c2a586a8c5abec1a76e305f28f8f0b48fe2fe27")
}

repo_raw_url <- function(ref, path) {
  paste0("https://raw.githubusercontent.com/PacificCommunity/",
         "ofp-sam-bet-2026-diagnostic/", ref, "/", path)
}

acquire_input <- function(name, remote_path, local_path = NULL) {
  if (!is.null(local_path) && nzchar(local_path)) {
    if (!file.exists(local_path)) stop("Local override does not exist: ", local_path)
    return(list(path = normalizePath(local_path), source = "local override"))
  }
  if (!identical(params$source_mode, "remote_cache")) {
    stop("source_mode must be 'remote_cache' unless an explicit local path is supplied")
  }

  cache_path <- file.path(params$cache_dir, params$source_ref, name)
  if (isTRUE(params$refresh_inputs) || !file.exists(cache_path)) {
    dir.create(dirname(cache_path), recursive = TRUE, showWarnings = FALSE)
    temporary_path <- paste0(cache_path, ".download")
    unlink(temporary_path)
    status <- tryCatch(
      download.file(repo_raw_url(params$source_ref, remote_path), temporary_path,
                    mode = "wb", quiet = TRUE),
      error = function(error) error
    )
    if (inherits(status, "error") || !identical(status, 0L) ||
        !file.exists(temporary_path) || file.info(temporary_path)$size == 0) {
      unlink(temporary_path)
      stop("Could not download ", remote_path, " at ref ", params$source_ref,
           ". Re-render after restoring network access or provide an explicit local path.")
    }
    backup_path <- paste0(cache_path, ".previous")
    unlink(backup_path)
    if (file.exists(cache_path) && !file.rename(cache_path, backup_path)) {
      unlink(temporary_path)
      stop("Could not preserve existing cached input: ", cache_path)
    }
    if (!file.rename(temporary_path, cache_path)) {
      if (file.exists(backup_path)) file.rename(backup_path, cache_path)
      unlink(temporary_path)
      stop("Could not move downloaded input into cache: ", cache_path)
    }
    unlink(backup_path)
    return(list(path = normalizePath(cache_path), source = "downloaded"))
  }
  list(path = normalizePath(cache_path), source = "cache")
}

input_files <- list(
  tag = acquire_input("bet.tag", "model/bet.tag", params$tag_path),
  par = acquire_input("final.par", "results/reference/final.par", params$par_path)
)
stopifnot(file.exists(input_files$tag$path), file.exists(input_files$par$path))

# Formatting helpers, matching the conventions of the companion simulation study.
fpc  <- function(x, d = 1) sprintf(paste0("%.", d, "f%%"), 100 * x)
fnum <- function(x) formatC(x, big.mark = ",", format = "d")
```

::: {.callout-note}
## Scope

This note is an executive summary of a methodological problem in the 2026 bigeye
diagnostic model: the treatment of tag recaptures during the mixing period, controlled
by `tag_flags(i,2)`. It computes, directly from `bet.tag` and the model's flag settings,
the two quantities that govern the problem — the share of recapture data excluded from
the likelihood, and the mixing-period recapture fraction $\hat\psi$ raised by the fitted
fishery-specific reporting rates — and reads them against a companion simulation study
(*Mixing-period tag recaptures in MULTIFAN-CL*, v10) whose results are quoted with
provenance, not re-derived.

Everything computed here is parsed inline from the `.tag` and fitted `.par` inputs;
everything quoted from the simulation study or reported from the assessment's
uncertainty grid is labelled as such. The recommendation at the end — a mixing-period
likelihood as the resolution, and a tag-free fleets-as-areas configuration as a candidate
structural alternative — follows from the demonstrated deficiencies of both flag
settings. Appendix A1 of the 2026 bigeye full paper reports that the already-developed
fleets-as-areas model gives comparable depletion and stock-status results; that result
is cited here, not reproduced. No document in this series simulates a fleets-as-areas
configuration.
:::

::: {.callout-note}
## Input provenance

Model inputs are pinned to
[`ofp-sam-bet-2026-diagnostic`](https://github.com/PacificCommunity/ofp-sam-bet-2026-diagnostic)
commit `r params$source_ref`. The tag file was resolved from
`r input_files$tag$source` at `r input_files$tag$path`; the fitted parameter file was
resolved from `r input_files$par$source` at `r input_files$par$path`.

When no local overrides are supplied, the document reads
`model/bet.tag` and `results/reference/final.par` from the pinned GitHub revision and
caches them under `r params$cache_dir`.
:::

::: {.callout-warning}
## Disclaimer: use of AI assistance

This note was prepared with the assistance of large language models.
The model parsed `bet.tag`, transcribed the flag settings and the `tag fish rep`
reporting-rate matrix from the supplied diagnostic `.par`, computed the diagnostics, and
drafted the text.

Output produced this way requires independent verification, and the document is
structured so that verification is possible rather than requiring trust:

- all computation is shown inline and runs in base R with no package dependencies;
- structural tripwires (`stopifnot`) check the parse against the file's own recovery
  header before any result is used, so a parsing failure halts the render rather than
  producing wrong numbers;
- the R implementation mirrors an independently written and executed implementation;
  any divergence between the two will surface as a tripwire failure or as numbers
  differing from the companion diagnostic note (v2), which reports the same quantities;
- the per-group appendix permits any single row to be checked by hand against the
  `.tag` file in a few minutes;
- every value quoted from the simulation study is attributed to it, and every value
  reported from the assessment's uncertainty grid is labelled as reported, not verified.

The inputs are read from a pinned upstream model revision or explicit local overrides.
The parser's section and dimension tripwires should still be checked before relying on
the document. Conclusions and recommendations are the responsibility of the author,
not of the tool.
:::

```{r constants}
#| code-summary: "Parse fitted tag settings and define quoted comparison inputs"

read_numeric_section <- function(path, marker, n_columns) {
  lines <- gsub("\r$", "", readLines(path, warn = FALSE))
  marker_at <- grep(paste0("^#\\s*", marker, "\\s*$"), lines,
                    ignore.case = TRUE)
  if (length(marker_at) != 1) stop("Expected one '# ", marker, "' section in ", path)
  next_marker <- grep("^#", lines[(marker_at + 1):length(lines)])[1]
  end_at <- if (is.na(next_marker)) length(lines) else marker_at + next_marker - 1
  rows <- trimws(lines[(marker_at + 1):end_at])
  rows <- rows[nzchar(rows)]
  fields <- strsplit(rows, "\\s+")
  if (!length(fields) || any(lengths(fields) != n_columns)) {
    stop("Malformed '# ", marker, "' section in ", path)
  }
  values <- suppressWarnings(as.numeric(unlist(fields, use.names = FALSE)))
  if (anyNA(values)) stop("Non-numeric value in '# ", marker, "' section in ", path)
  matrix(values, ncol = n_columns, byrow = TRUE)
}

tag_flags <- read_numeric_section(input_files$par$path, "tag flags", 10)
RR <- read_numeric_section(input_files$par$path, "tag fish rep", 33)
nmix <- tag_flags[, 1]

stopifnot(
  nrow(tag_flags) == 98,
  nrow(RR) >= 98,
  length(nmix) == 98,
  all(tag_flags[, 2] == 1),
  ncol(RR) == 33
)

RR <- RR[seq_len(98), , drop = FALSE]

# Regional seeding-prior means (SC22-SA-IP-05, purse seine), used only for comparison.
X_reg <- c(`1` = 0.5, `2` = 0.496, `3` = 0.512, `4` = 0.528, `5` = 0.5)

# --- Quoted from the companion simulation study (v11 fitted objects), not re-derived. ---
sim <- list(
  expD = data.frame(psi  = c(0.05, 0.30, 0.50),                    # Experiment D:
                    best = c(-0.039, -0.277, -0.509),              # bias in mixing-period F
                    worst= c(-0.048, -0.334, -0.599)),             # worst end = longest mixing
  floor_tau   = 0.22,     # survival floor; engages when removal fraction > 1 - tau = 0.78
  expE_N      = 125,      # Experiment E small-group design point
  expE_tail   = 0.060,    # share of groups with F-hat > 5x truth at that size
  expE_tail_big = 0.000,  # the same at 10,000 tags per group
  flag_gap_max  = 0.63)   # largest flag0-vs-truth gap in design (phi = 2, nmix = 4)
```

```{r parse}
#| code-summary: "Parse bet.tag (base R) with structural tripwires"

lines <- gsub("\r$", "", readLines(input_files$tag$path))

hdr <- as.integer(strsplit(trimws(lines[2]), "\\s+")[[1]])
n_groups <- hdr[1]; n_bins <- hdr[3]
stopifnot(n_groups == 98)

rec_hdr_at <- grep("^# TAG RECOVERIES", lines)[1]
totals <- as.integer(strsplit(trimws(lines[rec_hdr_at + 2]), "\\s+")[[1]])
stopifnot(length(totals) == n_groups)

blk <- grep("^#\\s+[0-9]+\\s+-\\s+RELEASE REGION", lines)
stopifnot(length(blk) == n_groups)
ends <- c(blk[-1] - 1, length(lines))

groups <- vector("list", n_groups)
recs   <- vector("list", n_groups)
for (g in seq_len(n_groups)) {
  seg  <- lines[blk[g]:ends[g]]
  prog <- tail(strsplit(trimws(seg[1]), "\\s+")[[1]], 1)
  body <- seg[-1]
  keep <- !grepl("^\\s*#", body) & nzchar(trimws(body))
  body <- trimws(body[keep])
  relh <- as.integer(strsplit(body[1], "\\s+")[[1]])          # region, year, month
  rel  <- as.numeric(strsplit(body[2], "\\s+")[[1]])
  stopifnot(length(rel) == n_bins)
  rmat <- if (length(body) > 2) {
    do.call(rbind, lapply(strsplit(body[-(1:2)], "\\s+"), as.integer))
  } else matrix(integer(0), 0, 5)
  stopifnot(nrow(rmat) == totals[g])   # header row counts recovery records per group
  q <- if (nrow(rmat)) ((rmat[, 3] - relh[2]) * 12 + (rmat[, 4] - relh[3])) %/% 3L
       else integer(0)
  stopifnot(all(q >= 0), all(((rmat[, 3] - relh[2]) * 12 + (rmat[, 4] - relh[3])) %% 3 == 0))
  groups[[g]] <- data.frame(gid = g, prog = prog, region = relh[1], year = relh[2],
                            N = sum(rel))
  recs[[g]]   <- if (nrow(rmat)) data.frame(gid = g, q = q, fishery = rmat[, 2],
                                            n = rmat[, 5]) else NULL
}
grp <- do.call(rbind, groups)
rec <- do.call(rbind, recs)
stopifnot(identical(grp$prog, c(rep("RTTP", 15), rep("PTTP", 46), rep("JPTP", 37))))
stopifnot(n_groups == nrow(tag_flags), all(rec$fishery >= 1),
          all(rec$fishery <= ncol(RR)))

grp$nmix <- nmix
rec$rate <- RR[cbind(rec$gid, rec$fishery)]
rec$mix  <- rec$q < grp$nmix[rec$gid]

N_tot <- sum(grp$N); R_tot <- sum(rec$n)
# Note: the recovery header row counts records, not tags; it is checked per group above.
```

```{r compute}
#| code-summary: "The two governing quantities"

# 1. Share of recapture data excluded from the likelihood by the mixing windows.
R_mix   <- sum(rec$n[rec$mix])
mix_share <- R_mix / R_tot

# 2. psi-hat per group: mixing returns raised by the fitted rate of the reporting
#    fishery (zero-rate returns flagged and excluded), divided by adjusted releases.
zero_mix <- sum(rec$n[rec$mix & rec$rate == 0])
raised <- with(rec[rec$mix & rec$rate > 0, ], tapply(n / rate, gid, sum))
grp$Rmix <- as.integer(with(rec[rec$mix, , drop = FALSE],
                            tapply(n, factor(gid, levels = 1:98), sum)))
grp$Rmix[is.na(grp$Rmix)] <- 0L
grp$psi <- 0; grp$psi[as.integer(names(raised))] <- raised
grp$psi <- grp$psi / grp$N
grp$psi_reg <- grp$Rmix / (X_reg[as.character(grp$region)] * grp$N)
grp$R <- with(rec, tapply(n, factor(gid, levels = 1:98), sum)); grp$R[is.na(grp$R)] <- 0

w_psi     <- sum(pmin(grp$psi, 1)     * grp$N) / N_tot
w_psi_reg <- sum(pmin(grp$psi_reg, 1) * grp$N) / N_tot
w_by_prog <- sapply(split(grp, grp$prog), function(d)
  sum(pmin(d$psi, 1) * d$N) / sum(d$N))

hi <- grp$psi >= 0.30
rec_hi <- sum(grp$R[hi]) / R_tot

# Switching flag=1 -> flag=0: extra raised removals by rate band.
band <- cut(rec$rate, c(-1, 0, 0.4, 0.95, 1.01),
            labels = c("zero-rate", "longline 0.16-0.26", "purse seine 0.46-0.84",
                       "JP coastal 0.99"))
extra <- with(rec[rec$mix & rec$rate > 0, ],
              tapply(n / rate - n, band[rec$mix & rec$rate > 0], sum))
extra[is.na(extra)] <- 0

# Counterfactual flag=0 floor margins under fitted rates and configured windows.
floor_util <- sapply(seq_len(98), function(g) {
  d <- rec[rec$gid == g & rec$mix & rec$rate > 0, , drop = FALSE]
  N <- grp$N[g]; u <- 0
  for (t in 0:(max(grp$nmix[g], 1) - 1)) {
    rt <- sum(d$n[d$q == t] / d$rate[d$q == t])
    if (N <= 0) return(Inf)
    u <- max(u, (rt / N) / (1 - sim$floor_tau)); N <- N - rt
  }
  u
})
n_floor <- sum(floor_util >= 1)
grp$raised_mix <- grp$psi * grp$N
grp$unremoved_flag1 <- pmax(grp$raised_mix - grp$Rmix, 0)
grp$floor_util_flag0 <- floor_util
```

```{r minimal-simulations}
#| code-summary: "Minimal Experiment D calibration and forward_X proof of concept"

# This deliberately small operating model is an executable illustration of the
# mechanism, not a re-run of the companion study or an assessment operating model.
baranov <- function(f, n, m) {
  z <- f + m
  (f / z) * (1 - exp(-z)) * n
}

om <- list(N0 = 10000, Tmax = 8, M = 0.20, F_true = 0.15,
           X_assumed = 0.50, nmix = 2L, phi = 1)

psi_of <- function(p, phi, nmix) {
  n <- p$N0
  total <- 0
  for (t in seq_len(nmix)) {
    total <- total + baranov(p$F_true * phi, n, p$M)
    n <- n * exp(-(p$F_true * phi + p$M))
  }
  total / p$N0
}

psi_closed <- function(p, phi, nmix) {
  z <- p$F_true * phi + p$M
  (p$F_true * phi / z) * (1 - exp(-nmix * z))
}

phi_for_psi <- function(target, nmix, p = om, phi_max = 30) {
  if (target <= 0) return(0)
  if (psi_of(p, phi_max, nmix) < target) return(NA_real_)
  uniroot(function(phi) psi_of(p, phi, nmix) - target,
          interval = c(1e-8, phi_max), tol = 1e-10)$root
}

bias_pred <- function(p, phi, nmix, x) {
  psi <- psi_of(p, phi, nmix)
  n_post <- p$N0 * exp(-(p$F_true * phi + p$M) * nmix)
  unremoved <- (1 - x) * psi * p$N0
  -unremoved / (n_post + unremoved)
}

# Matched-psi Experiment D calibration: expected, not Monte Carlo, results.
expd <- expand.grid(psi_target = c(0.05, 0.30, 0.50), nmix = 2:4)
expd$phi <- mapply(phi_for_psi, expd$psi_target, expd$nmix)
expd$psi <- mapply(function(phi, nmix) psi_of(om, phi, nmix),
                   expd$phi, expd$nmix)
expd$bias <- mapply(function(phi, nmix) bias_pred(om, phi, nmix, om$X_assumed),
                    expd$phi, expd$nmix)
stopifnot(
  all(is.finite(expd$phi)),
  max(abs(expd$psi - expd$psi_target)) < 1e-8,
  all(expd$bias < 0),
  all(abs(mapply(function(phi, nmix) psi_of(om, phi, nmix) -
                   psi_closed(om, phi, nmix),
                 expd$phi, expd$nmix)) < 1e-12)
)

# Conditional application of that approximation to this file. Each group is matched
# on its observed psi-hat and mixing length, using its effective fitted reporting rate.
eligible_reported <- with(rec[rec$mix & rec$rate > 0, ],
                          tapply(n, factor(gid, levels = 1:98), sum))
eligible_reported[is.na(eligible_reported)] <- 0
grp$x_effective <- ifelse(grp$raised_mix > 0, eligible_reported / grp$raised_mix, NA)
grp$phi_matched <- mapply(phi_for_psi, grp$psi, grp$nmix)
grp$flag1_bias_illustrative <- mapply(
  function(phi, nmix, x) {
    if (!is.finite(phi) || !is.finite(x) || x <= 0 || x > 1) return(NA_real_)
    bias_pred(om, phi, nmix, x)
  }, grp$phi_matched, grp$nmix, grp$x_effective)

# Minimal stochastic forward_X proof of concept.
posfun <- function(x, eps) {
  if (x >= eps) return(c(y = x, pen = 0))
  c(y = eps / (2 - x / eps), pen = (x - eps)^2)
}

solve_mixing_f <- function(target, n, m, iter = 25) {
  if (target <= 0) return(0)
  f <- -log(1 - min(target / n, 0.999))
  for (iteration in seq_len(iter)) {
    z <- f + m
    e <- exp(-z)
    gradient <- (f / z) * (1 - e) * n - target
    derivative <- n * ((m * (1 - e)) / z^2 + (f / z) * e)
    f <- max(f - gradient / derivative, 1e-10)
  }
  f
}

mix_step <- function(r, n, p, mode, floor = sim$floor_tau) {
  target <- if (mode == "flag0") r / p$X_assumed else r
  survival <- 1 - target / n
  floored <- survival < floor
  penalty <- 0
  if (floored) {
    adjusted <- posfun(survival - 0.02, floor - 0.02)
    survival <- 0.02 + adjusted[["y"]]
    penalty <- adjusted[["pen"]]
    target <- n * (1 - survival)
  }
  list(f = solve_mixing_f(target, n, p$M), penalty = penalty, floored = floored)
}

simulate_returns <- function(p, x_true) {
  n <- p$N0
  returns <- numeric(p$Tmax)
  for (t in seq_len(p$Tmax)) {
    f <- p$F_true * if (t <= p$nmix) p$phi else 1
    returns[t] <- rpois(1, x_true * baranov(f, n, p$M))
    n <- n * exp(-(f + p$M))
  }
  returns
}

nll_tag <- function(par, returns, p, mode, prior_sd = 0.25) {
  f_pop <- exp(par[1])
  phi <- if (mode == "forward_X") exp(par[2]) else 1
  x <- if (mode == "forward_X") plogis(par[3]) else p$X_assumed
  n <- p$N0
  nll <- 0
  for (t in seq_len(p$Tmax)) {
    if (t <= p$nmix && mode %in% c("flag0", "flag1")) {
      step <- mix_step(returns[t], n, p, mode)
      f <- step$f
      nll <- nll + 1000 * step$penalty
    } else {
      f <- if (t <= p$nmix) f_pop * phi else f_pop
      mu <- x * baranov(f, n, p$M) + 1e-10
      nll <- nll - dpois(returns[t], mu, log = TRUE)
    }
    n <- n * exp(-(f + p$M))
  }
  if (mode == "forward_X") {
    nll <- nll + 0.5 * ((log(x) - log(p$X_assumed)) / prior_sd)^2
  }
  nll
}

fit_one <- function(returns, p, mode) {
  if (mode %in% c("flag0", "flag1")) {
    interval <- log(c(1e-3, 5))
    fit <- optimize(function(log_f) nll_tag(c(log_f, 0, 0), returns, p, mode),
                    interval = interval, tol = 1e-9)
    return(c(F_hat = exp(fit$minimum), X_hat = p$X_assumed, conv = 0,
             at_bound = as.numeric(any(abs(fit$minimum - interval) < 1e-3))))
  }
  fit <- optim(c(log(0.10), log(1), qlogis(p$X_assumed)),
               function(par) nll_tag(par, returns, p, mode), method = "Nelder-Mead",
               control = list(maxit = 3000, reltol = 1e-10))
  c(F_hat = exp(fit$par[1]), X_hat = plogis(fit$par[3]), conv = fit$convergence,
    at_bound = 0)
}

set.seed(20260810)
n_sim <- 100
forward_design <- expand.grid(x_true = c(0.35, 0.50, 0.65),
                              mode = c("flag0", "flag1", "forward_X"),
                              replicate = seq_len(n_sim), KEEP.OUT.ATTRS = FALSE)
forward_res <- do.call(rbind, lapply(seq_len(nrow(forward_design)), function(row) {
  setting <- forward_design[row, ]
  fit <- fit_one(simulate_returns(om, setting$x_true), om, setting$mode)
  data.frame(x_true = setting$x_true, mode = setting$mode,
             F_hat = fit[["F_hat"]], X_hat = fit[["X_hat"]],
             conv = fit[["conv"]], at_bound = fit[["at_bound"]])
}))
forward_res$rel_bias <- forward_res$F_hat / om$F_true - 1
forward_summary <- do.call(rbind, lapply(split(forward_res,
                                                list(forward_res$x_true,
                                                     forward_res$mode), drop = TRUE),
  function(d) data.frame(`True reporting rate` = d$x_true[1], Treatment = d$mode[1],
                         `Median F bias` = median(d$rel_bias),
                         `Forward_X convergence` = mean(d$conv == 0),
                         `Flag endpoint contact` = mean(d$at_bound == 1),
                         check.names = FALSE)))
forward_summary <- forward_summary[order(forward_summary[[1]], forward_summary[[2]]), ]
stopifnot(!any(forward_res$at_bound[forward_res$mode %in% c("flag0", "flag1")] == 1))
```

# How MULTIFAN-CL treats mixing returns {#sec-mechanics}

The issue is not that MULTIFAN-CL lacks a tag likelihood. **After mixing**, it projects
the tagged population with the population fishing mortality, predicts reported returns
as expected recaptures multiplied by the reporting rate, and scores those returns in the
tag likelihood. **During mixing**, it instead uses observed returns to solve
deterministically for a tag-specific removal that fixes the tagged population state.
The mixing returns themselves contribute no likelihood term under either flag setting.

| `tag_flags(i,2)` | Mixing-period removal target | Mixing returns scored in likelihood? | Principal failure mode |
|---:|---|:---:|---|
| 0 | $R_{obs} / X$ | No | Conditionally valid only if the assumed reporting rate is right, the survival floor is inactive, and the release-group inversion is informative. |
| 1 | $R_{obs}$ | No | Assumes complete reporting and systematically under-removes wherever $X < 1$. |

Here $R_{obs}$ is the reported return and $X$ is its reporting rate. The first setting
raises observed returns for non-reporting before removal; the second treats reported
returns as true recaptures. Neither treatment lets the model estimate a disagreement
between predicted and observed mixing returns, attach uncertainty to them, or update
parameters through a mixing-period likelihood. This is why choosing a flag is not a
likelihood-based treatment of the data.

## Why `tag_flags(i,2) = 1` exists {#sec-floor}

`tag_flags(i,2) = 1` exists to provide an escape from a failure mode of the original,
rate-raised treatment. Under `tag_flags(i,2) = 0`, MULTIFAN-CL divides a reported
mixing-period return by its assumed reporting rate before removing tags. This is the
natural correction for incomplete reporting, but it can become numerically and
biologically implausible when the raised removal is large relative to the tagged cohort.

The survival-floor pathology is what happens next. With a low reporting rate, a large
observed return, or both, the implied removal in one mixing period can leave fewer tags
than MULTIFAN-CL permits to survive. The model then applies `posfun`: it floors survival
and rescales removal to the amount consistent with that floor. Rather than removing
$R_{obs}/X$, it removes less. The raised treatment therefore becomes an under-removal
treatment at precisely the point it was intended to correct under-reporting, pulling
inferred $F$ downward and biomass upward.

Flag 1 avoids that rate-raised floor pathway by removing only $R_{obs}$. It does not
solve the statistical problem. It trades a conditional cap pathology for the
unconditional assumption that reporting is complete, so it under-removes whenever
$X < 1$, whether or not the floor would have engaged under flag 0.

For this file, the fitted fishery-specific reporting rates and configured mixing windows
screen `r n_floor` group(s) as reaching the configured floor under flag 0, rather than
the six groups suggested by regional-mean raising. This makes the floor a targeted
flag-0 fit diagnostic, not a general reason to select flag 1 and not evidence that it is
absent. The current rates were fitted under flag 1 and would re-equilibrate under flag 0;
an actual flag-0 fit must identify any floor engagements and their objective-function
contribution.

# Summary {#sec-summary}

**Two structural facts frame everything.** First, under the configured mixing windows,
`r fnum(R_mix)` of the `r fnum(R_tot)` reported recaptures in `bet.tag` —
**`r fpc(mix_share)` of the tagging data** — never enter the likelihood. Under either
setting of `tag_flags(i,2)`, those returns fix the tagged population state
deterministically, with full authority and no uncertainty attached, and contribute
nothing to parameter estimation. Second, the removal that fixes that state depends
entirely on what is assumed about reporting rates during mixing, and the model's own
fitted `tag fish rep` matrix says those rates are far from uniform: near-complete
(0.99) for the Japanese coastal fisheries that dominate JPTP returns, roughly one half
(0.46--0.50) for the main purse-seine fisheries that dominate PTTP and RTTP returns,
and 0.16--0.26 for longline.

**The configured setting, `tag_flags(i,2) = 1`, removes reported mixing-period returns
at face value — equivalent to assuming complete reporting — and its cost falls almost
entirely on the purse-seine programmes.** By the model's own fitted rates the
assumption is nearly true for JPTP (weighted $\hat\psi$ = `r round(w_by_prog["JPTP"], 3)`;
`r fpc(sum(rec$n[rec$mix & rec$fishery %in% c(12,13) & grp$prog[rec$gid] == "JPTP"]) / sum(rec$n[rec$mix & grp$prog[rec$gid] == "JPTP"]), 0)`
of JPTP mixing returns come from the 0.99-rate fisheries) and substantially false for
RTTP (`r round(w_by_prog["RTTP"], 3)`) and PTTP (`r round(w_by_prog["PTTP"], 3)`), where
roughly a third to a half of true mixing recaptures are never removed. The companion
study's Experiment D maps $\hat\psi$ to the downward bias in mixing-period $F$:
`r fpc(abs(sim$expD$best[1]), 1)`--`r fpc(abs(sim$expD$worst[1]), 1)` at $\psi$ = 0.05,
`r fpc(abs(sim$expD$best[2]), 1)`--`r fpc(abs(sim$expD$worst[2]), 1)` at 0.3, and
`r fpc(abs(sim$expD$best[3]), 1)`--`r fpc(abs(sim$expD$worst[3]), 1)` at 0.5, the worse
end at longer mixing. `r sum(hi)` release groups sit at $\hat\psi \ge 0.3$, carrying
`r fpc(rec_hi, 0)` of the reported recaptures. The bias is systematic — $F$ low, scale
high, every group the same direction — and its one merit is that it is predictable in
advance from exactly this calculation.

**The alternative, `tag_flags(i,2) = 0`, raises returns by the assumed rate before
removal, and is unbiased only under conditions this file does not fully meet.** Its
accuracy moves one-for-one with the reporting rates, which are the quantity under
dispute, weakly informed for longline, and being revised downward. Its inversion is
convex in a noisy observation, so small release groups put weight in the upper tail of
$F$: at the study's `r sim$expE_N`-tag design point, `r fpc(sim$expE_tail, 1)` of
simulated groups returned estimates more than five times the truth with everything else
correct, invisible to median-based diagnostics — and `r sum(grp$N <= sim$expE_N)` of the
98 groups here are that small or smaller. The oft-cited floor pathology, by contrast,
largely dissolves under the fitted rates: re-evaluated with fishery-specific raising and
the configured windows, `r n_floor` group(s) breach the survival-floor condition, not
the six that regional-mean raising suggests. And the switch is surgically targeted:
moving to raised removals adds `r fnum(round(sum(extra[c("purse seine 0.46-0.84","longline 0.16-0.26")])))`
tags to the purse-seine and longline removals — where the face-value bias lives — while
changing JPTP by `r fnum(round(extra["JP coastal 0.99"]))` tags in total.

**Running both settings as levels of the uncertainty grid does not bound the problem.**
The pair is structurally one-sided: because the assumed rate is below one wherever it
matters, the raised treatment always removes more tags and always returns the larger
$F$. Two grid levels therefore span an assumption without bracketing the truth — they
cannot contain it at all when the assumed rate is too high — and, because one level
carries a documented directional bias, averaging them functions as a fifty per cent
weighting on that bias, not as an uncertainty representation.

**This is not academic: the choice reportedly flips the status determination.** With
both settings in the grid, the ensemble indicates the stock is not overfished and not
undergoing overfishing; restricted to `tag_flags(i,2) = 0`, it indicates overfished and
about as likely as not undergoing overfishing. Those ensemble results are reported here,
not verified, but their direction is exactly what the $\hat\psi$ analysis predicts, and
the guidance against the grid axis was fixed before the ensemble outcome was known.
The honest statement of status is therefore not either determination — it is that the
determination currently turns on a model component that is biased under both of its
available settings.

# Assessment guidance {#sec-guidance}

## First-level guidance: resolve or replace the tagged formulation

**The resolution is a mixing-period likelihood.** During mixing, MULTIFAN-CL should
project the tagged cohort, predict reported returns as $X C$, score those returns against
the observations, and estimate reporting rates against their tag-seeding priors. This
would put `r fpc(mix_share, 0)` of the tag data back into estimation and remove the
deterministic inversion. The companion simulation's `forward_X` treatment recovers $F$
under reporting-rate misspecification that biases both flags, but it is not free: when
release groups are small, estimating fishing mortality, mixing availability, and
reporting rates together can be imprecise. This is a MULTIFAN-CL redevelopment and
validation task, not a configuration change for the current assessment.

Formally, for release group $i$, period $t$, and fishery $f$, `forward_X` advances the
tagged cohort $N_{it}$ using a mixing-period fishing mortality $F^T_{it}$ and natural
mortality $M$, forms true recaptures $C_{itf}$ from that state, and predicts

$$
\mu_{itf} = X_{if} C_{itf}.
$$

The mixing return then contributes an observation likelihood, for example
$R_{itf} \sim \operatorname{Poisson}(\mu_{itf})$, instead of a deterministic removal
target. A tag-seeding prior supplies information on the reporting rate, such as a
penalty on $\log X_{if}$ around its prior mean. In the proof of concept, the free
parameters are population fishing mortality ($\log F$), a mixing-period availability
multiplier ($\log \varphi$), and reporting rate ($\operatorname{logit}X$). The flags
estimate none of the latter two from mixing returns; they solve for whatever
mixing-period removal reproduces their chosen target. `forward_X` consequently lets the
data, process, and prior disagree in an estimable way, but it can be weakly identified
when many small release groups must inform all three quantities.

**A tag-free fleets-as-areas model is also a credible path for model development and
management advice.** Appendix A1 of the
[2026 bigeye full paper](https://meetings.wcpfc.int/node/32657) reports that the
already-developed, spatially implicit fleets-as-areas model has comparable depletion and
stock-status results to the current tagged model. That agreement is reported here, not
reproduced. Subject to its own fit and diagnostic evaluation, this is a candidate full
assessment model and may be included in the structural uncertainty grid. It represents
uncertainty in the combined data-and-spatial formulation, not a correction for the tag
flags or an estimate of their isolated effect. It does not validate the tagged model's
mixing-period treatment.

## Second-level guidance: the tagged spatial branch

For this tag file, `tag_flags(i,2) = 1` should not be retained as the selected tagged
configuration. It removes reported returns at face value even though the mixing returns
that matter most are not fully reported: weighted $\hat\psi$ is
`r round(w_by_prog["RTTP"], 3)` for RTTP and `r round(w_by_prog["PTTP"], 3)` for PTTP,
compared with `r round(w_by_prog["JPTP"], 3)` for JPTP. `r sum(hi)` groups with
$\hat\psi \ge 0.3$ carry `r fpc(rec_hi, 0)` of all reported recaptures. The configured
face-value treatment therefore imposes its known directional bias on a material share of
the data, concentrated in programmes with purse-seine reporting rates near one half.
It can remain a documented diagnostic sensitivity, but not an equally weighted grid level
or a bounding endpoint.

If a tagged spatial branch must be retained before either first-level path is available,
`tag_flags(i,2) = 0` is the provisionally preferable single setting, with three explicit
conditions attached to this file:

1. **Confirm floor behaviour in an actual flag-0 fit.** The fitted-rate screen identifies
  `r n_floor` group(s) whose raised mixing removal reaches the configured survival floor.
  That is substantially less than the regional-mean screen suggests, but it is not a
  clean pass: fitted reporting rates would re-equilibrate after the flag changes. The
  actual fit must report the affected group(s), the `posfun` contribution, and the
  resulting floor utilization before its outputs are used for advice.
2. **Treat small-group inversion as a tail diagnostic, not a median diagnostic.**
  `r sum(grp$N <= sim$expE_N)` of the 98 release groups contain `r sim$expE_N` tags or
  fewer. The simulation result implies that their raised-return inversions can produce
  an upper tail in $F$ even with correct reporting rates and an inactive floor. The
  flag-0 run should therefore report group-level implied mixing mortality and any
  optimizer-bound or extreme-return diagnostics, rather than relying on a pooled median
  or likelihood summary.
3. **Represent reporting-rate uncertainty directly.** Switching to flag 0 raises
  `r fnum(round(sum(extra[c("purse seine 0.46-0.84", "longline 0.16-0.26")])))` mixing
  returns in the purse-seine and longline rate bands, but only
  `r fnum(round(extra["JP coastal 0.99"]))` in the near-complete-reporting Japanese
  coastal band. The change is therefore driven by precisely the reporting rates under
  active scrutiny. Sensitivity to their priors or fixed assumptions belongs in the
  tagged-model diagnostic set; replacing it with a flag-1 grid level would hide rather
  than represent that uncertainty.

The `r fnum(zero_mix)` zero-rate mixing returns are unresolved under either choice and
should be traced separately. Neither setting converts the excluded
`r fpc(mix_share, 0)` of recaptures into probabilistic evidence, so this provisional
choice does not remove the case for the likelihood redevelopment or the tag-free
fleets-as-areas alternative.

# The $\hat\psi$ table {#sec-psi}

For release group $i$, the empirical diagnostic in this note is

$$
\hat\psi_i = \frac{1}{N_i}\sum_{t < n_{mix,i}}\sum_f
\frac{R_{itf}}{\hat X_{if}},
$$

where $R_{itf}$ is a reported mixing-window return from fishery $f$, $\hat X_{if}$ is
its fitted reporting rate, and $N_i$ is the correction-adjusted release count. Thus
$\hat\psi_i$ estimates the fraction of the group truly recaptured while the mixing rule
is active; it is not the fraction merely reported. Omitting the raising returns
$X\psi$ rather than $\psi$ — roughly half the diagnosed fraction at purse-seine rates,
and because the bias steepens in $\psi$, less than half the diagnosed consequence.

The companion simulation explains why this quantity governs `tag_flags(i,2) = 1`.
That setting removes $R$ where it should remove the true recaptures $C$, leaving roughly
$(1-X)\psi N_0$ extra tags at liberty. In the stripped-down operating model this gives
the approximation

$$
\frac{\hat F}{F} \approx
\frac{N_{post}}{N_{post} + (1-X)\psi N_0},
$$

so the effect is systematically toward lower $F$ and higher biomass. This is an
analytical interpretation of the simulation, not a fitted assessment prediction.
`r fnum(zero_mix)` mixing returns come from a fishery whose fitted rate is zero and
cannot be raised at all; they are excluded, which understates $\hat\psi$ for the
affected groups, and the inconsistency should be resolved regardless of the flag choice.

```{r psi-table}
#| echo: false
cls <- cut(grp$psi, c(-1, 0.05, 0.30, 0.50, Inf),
           labels = c("< 0.05", "0.05\u20130.30", "0.30\u20130.50", "\u2265 0.50"))
tab <- data.frame(
  `psi class` = levels(cls),
  Groups = as.integer(table(cls)),
  `Share of recaptures` = sapply(levels(cls), function(k)
    fpc(sum(grp$R[cls == k]) / R_tot, 1)),
  `Experiment D bias class (mixing-period F)` =
    c("a few per cent", "up to roughly one third",
      "roughly one third to one half", "one half or more"),
  check.names = FALSE)
knitr::kable(tab, row.names = FALSE)

byprog <- do.call(rbind, lapply(split(grp, grp$prog), function(d) data.frame(
  Programme = d$prog[1], Groups = nrow(d), Releases = fnum(round(sum(d$N))),
  `Mixing returns` = fnum(sum(d$Rmix)),
  `Weighted psi (fishery rates)` = round(sum(pmin(d$psi, 1) * d$N) / sum(d$N), 3),
  `(regional means)` = round(sum(pmin(d$psi_reg, 1) * d$N) / sum(d$N), 3),
  check.names = FALSE)))
knitr::kable(byprog[order(match(byprog$Programme, c("RTTP", "PTTP", "JPTP"))), ],
             row.names = FALSE)
```

The programme contrast carries the story: the face-value assumption is nearly correct
where the fitted rates are near one, and costs a third to a half of the true mixing
removals where they are near one half — which is where
`r fpc(sum(grp$N[grp$prog != "JPTP"]) / N_tot, 0)` of the releases sit.

## Conditional implication for this tag file

The following envelope applies the stripped-down Experiment D approximation to each
release group's observed $\hat\psi$, mixing length, and effective fitted reporting rate.
It is not an assessment estimate of $F$ bias: it holds the operating model's natural
mortality and post-mixing fishing mortality fixed, infers the availability needed to
produce the observed $\hat\psi$, and then asks what `tag_flags(i,2) = 1` would imply in
that simplified setting. Its purpose is to make the direction and potential scale of the
known face-value bias explicit for this file.

```{r observed-psi-envelope}
#| echo: false
envelope <- do.call(rbind, lapply(split(grp, grp$prog), function(d) {
  bias <- d$flag1_bias_illustrative[is.finite(d$flag1_bias_illustrative)]
  data.frame(
    Programme = d$prog[1],
    `Calibrated groups` = length(bias),
    `Weighted psi` = round(sum(pmin(d$psi, 1) * d$N) / sum(d$N), 3),
    `Illustrative F bias range` = if (length(bias)) {
      paste0(fpc(min(bias), 0), " to ", fpc(max(bias), 0))
    } else "not calibrated",
    check.names = FALSE)
}))
knitr::kable(envelope[order(match(envelope$Programme, c("RTTP", "PTTP", "JPTP"))), ],
             row.names = FALSE)
```

# Mechanism checks {#sec-mechanism-checks}

This table makes the deterministic difference between the two flags auditable from the
same inputs as the $\hat\psi$ calculation. `Raised mixing returns` is the removal target
under `tag_flags(i,2) = 0`; `reported mixing returns` is the target under
`tag_flags(i,2) = 1`; their difference is the mixing-period removal omitted by the
face-value treatment. `Flag-0 floor utilization` is the largest quarterly raised removal
as a share of the amount that can be removed before the configured survival floor would
engage. Values at or above one require confirmation from an actual flag-0 fit because
reporting rates would re-equilibrate.

```{r mechanism-checks}
#| echo: false
mech <- data.frame(
  Group = grp$gid,
  Prog. = grp$prog,
  N = round(grp$N),
  `Reported mixing returns` = grp$Rmix,
  `Raised mixing returns` = round(grp$raised_mix),
  `Unremoved by flag 1` = round(grp$unremoved_flag1),
  `Flag-0 floor utilization` = round(grp$floor_util_flag0, 3),
  check.names = FALSE)
knitr::kable(mech, row.names = FALSE)
```

# Minimal simulation checks {#sec-minimal-simulations}

The companion study remains the full simulation evidence. The two checks below are kept
small enough to render with this note and make its two central claims reproducible: the
`flag=1` bias grows with the fraction recaptured while mixing; and a likelihood that
predicts rather than inverts mixing returns can recover from reporting-rate
misspecification. They are intentionally not an assessment operating model: one fishery,
one release group, Poisson reporting, fixed natural mortality, and no movement or age
structure.

## Matched-$\psi$ Experiment D calibration

For each target $\psi$, the calculation solves for the mixing availability that gives
that fraction at each mixing length and evaluates the analytical `flag=1` bias at
$X=0.5$. The checks in the code require the iterative and closed-form $\psi$ values to
agree and every calculated bias to be negative. At a common $\psi$, longer mixing still
has a somewhat larger consequence because fewer tags remain at liberty after mixing.

```{r expd-minimal-table}
#| echo: false
expd_table <- expd[, c("psi_target", "nmix", "phi", "bias")]
names(expd_table) <- c("Target psi", "Mixing periods", "Matched availability",
                       "Illustrative F bias")
expd_table$`Target psi` <- round(expd_table$`Target psi`, 2)
expd_table$`Matched availability` <- round(expd_table$`Matched availability`, 2)
expd_table$`Illustrative F bias` <- fpc(expd_table$`Illustrative F bias`, 1)
knitr::kable(expd_table[order(expd_table$`Target psi`, expd_table$`Mixing periods`), ],
             row.names = FALSE)
```

## `forward_X` proof of concept

This stochastic check simulates `r n_sim` release histories for each true reporting rate,
holding the analyst's prior centre at $X=0.5$. It compares the two deterministic flags
with `forward_X`, which estimates $F$, mixing availability, and $X$ jointly under its
reporting-rate prior. The convergence column is deliberately reported rather than
silently filtering failed `forward_X` fits; endpoint contacts are similarly reported for
the deterministic inversions.

```{r forward-x-table}
#| echo: false
forward_display <- forward_summary
forward_display$`Median F bias` <- fpc(forward_display$`Median F bias`, 1)
forward_display$`Forward_X convergence` <- fpc(forward_display$`Forward_X convergence`, 0)
forward_display$`Flag endpoint contact` <- fpc(forward_display$`Flag endpoint contact`, 0)
knitr::kable(forward_display, row.names = FALSE)
```

# Limitations

1. **The transcribed settings are the load-bearing inputs.** `nmix` and the
   reporting-rate patterns must be verified against the final `.par` before this note
   is relied on; a render against different settings will silently produce different —
   correct — numbers.
2. **The fitted rates were estimated under `tag_flags(i,2) = 1`.** Using them to
   characterise the raised counterfactual is indicative; under the default the rates
   would re-equilibrate, and the feedback runs toward the floor. The one-group floor
   finding should be confirmed from an actual `tag_flags(i,2) = 0` fit before it is
   quoted without qualification.
3. **The Experiment D mapping is from a deliberately stripped-down operating model.**
   It gives the class of the bias, not its value in a full assessment; the companion
   study's limitations section applies in full.
4. **The ensemble status results are reported, not reproduced.** This note takes the
   direction and existence of the flip as given by the assessment team; nothing here
   recomputes grid outputs.
5. **$\hat\psi$ inherits the rates it is raised by**, in the direction that understates
   the problem if reporting estimates fall further, and the zero-rate fishery-27
   returns are excluded from raising.
6. **The fleets-as-areas result is cited, not reproduced.** The statement that its
  depletion and stock-status results are comparable to the current tagged model is
  reported from Appendix A1 of the 2026 bigeye full paper. This note does not reproduce
  that model, assess its fits, or simulate its spatial representation.
7. **The embedded simulations are proof-of-concept checks.** They are intentionally
  smaller and simpler than the companion study, and do not quantify this assessment's
  fitted $F$ bias or validate a MULTIFAN-CL implementation. Their reporting-rate prior,
  observation model, and single-group design are illustrative.

# Appendix: per-group $\hat\psi$ {#sec-appendix}

```{r appendix}
#| echo: false
app <- data.frame(Group = grp$gid, Prog. = grp$prog, Region = grp$region,
                  Year = grp$year, `tag_flags(i,1)` = grp$nmix,
                  N = round(grp$N), R = grp$R, `R mix` = grp$Rmix,
                  `psi (fishery rates)` = round(grp$psi, 3),
                  `psi (regional means)` = round(grp$psi_reg, 3),
                  check.names = FALSE)
knitr::kable(app, row.names = FALSE)
```
